Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-04-25 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
small chenges and edits to highligthed text

\ab{We will treat the modified theories of gravity as EFTs, and focus on
finite-size effects (see, e.g., Ref.~\cite{Sennett:2019bpc}). Thus, for each gravity theory we impose that
the fundamental length-scale $\ell \lesssim GM/c^2$, where $M$ is the mass of the BH. This implies
that observable deviations from GR present in those theories arise from modifications
to the Kerr geometry of each individual BH. Those finite-size effects can manifest themselves in
the QNMs of the remnant produced by the merger, but also in the GW phasing of the inspiral through
corrections to the GR spin-induced quadrupole and Love numbers. However, here we will not consider the latter,
instead, we will only study the impact of finite-size effects on the QNMs of the
remnant.}

@@

The theory admits as a solution, the garden-variety Schwarzschild BH of GR.
This is not the case when rotation is included and the Kerr metric is not a
solution of the theory~\cite{Jackiw:2003pm}.
%
These rotating BHs support a scalar field which fall off as
$r^{-2}$ asymptotically, to which we can associate a scalar dipole
charge~\cite{Yunes:2009hc,Konno:2009kg} and
%
the leading-order modification to the GW phase enters at 2PN~\cite{Yagi:2011xp}.
%
Deviations from GR at this PN order are constrained with present GW observations~\cite{LIGOScientific:2021sio}
only at the level of $\sim 50\%$ (see the constraint on the 2PN parameter $\varphi_4$ in Fig.~6 of
Ref.~\cite{LIGOScientific:2021sio}). \ab{So far, analyses that used only the inspiral portion of
the BBH GW signals were not able to set meaningful bounds on such deviation at 2PN order~\cite{Nair:2019iur,Perkins:2021mhb}.}
%
Nonetheless, the theory has been constrained in Ref.~\cite{Silva:2020acr}, which found
%
$\ell_{\rm dCS} \leqslant 8.5$~km,
%
by folding data from the X-ray observations of the isolated NS
PSR~J0030+0451~\cite{Lommen:2000yt,NANOGrav:2017wvv} by
NICER~\cite{Riley:2019yda,Miller:2019cac} and from the GW observation of the binary
NS GW170817~\cite{TheLIGOScientific:2017qsa,LIGOScientific:2018cki}, using
equation-of-state independent relations between the NS moment of inertia and tidal deformability~\cite{Yagi:2013bca,Yagi:2013awa,Gupta:2017vsl}.

@@

More specifically, we follow the notation of
Refs.~\cite{Cano:2020cao,Cano:2021myl} and consider up to dimension-eight
operators ($n=4$),
%
\begin{subequations}
\label{eq:action_mods}
\begin{align}
    L^{(6)} &= \lame R_{\mu\nu}{}^{\rho\sigma} R_{\rho\sigma}{}^{\gamma\delta} R_{\gamma \delta}{}^{\mu\nu}
    + \lamo R_{\mu\nu}{}^{\rho\sigma} R_{\rho\sigma}{}^{\gamma\delta} \tilde{R}_{\gamma \delta}{}^{\mu\nu}\,,
    \label{eq:s_6}
    \\
    L^{(8)} &= \varepsilon_{1} \mathcal{C}^2
    + \varepsilon_{2} \tilde{\mathcal{C}}^{2}
    + \varepsilon_{3} \mathcal{C} \tilde{\mathcal{C}}\,,
\label{eq:s_8}
\end{align}
\end{subequations}
%
where ${\cal C} = R_{\mu\nu\rho\sigma} R^{\mu\nu\rho\sigma}$,
$\tilde{\cal C} = R_{\mu\nu\gamma\delta} \epsilon^{\mu\nu}{}_{\rho\sigma} R^{\rho\sigma\gamma\delta}$,
and $(\lambda_{\rm o, e},\varepsilon_{i})$, $i=$~1, 2, 3, are dimensionless parameters.
%
Due to the large number of free parameters in this theory, we focus on a subset of the parameter space.
%
In particular, we consider dimension-six and dimension-eight operators separately.
%
In addition, in the dimension-six case we further assume that $\lame = \lamo = 1$,
leaving us with $\ell_{\rm cEFT}$ as our single free parameter.
%
Similarly, in the dimension-eight case, we set $\varepsilon_{1} = 1$ and $\varepsilon_{2} = \varepsilon_{3} = 0$,
as done in Ref.~\cite{Sennett:2019bpc}.
%
This leaves us with $\ell_{\rm qEFT}$ as our single free parameter.

\ab{For the EFT of GR with dimension-eight operators, Ref.~\cite{Sennett:2019bpc} focused on the
orbital effects (i.e., instead of finite-size effects) and performed
Bayesian model selection using the two lowest-mass BBHs events of the second-observig run of
the LIGO-Virgo Collaboration, notably GW151226 and GW170608. They found that the
data disfavor the appearance of new physics on distance scales around $\ell_{\rm qEFT} \sim 150$ km.}

>>>>

We will treat the modified theories of gravity as EFTs, and focus on
finite-size effects (see, e.g., Ref.~\cite{Sennett:2019bpc}). Thus, for each gravity theory we impose that
the fundamental length-scale $\ell_{\rm th} \lesssim GM/c^2$, where $M$ is the mass of the BH. This implies
that observable deviations from GR present in those theories arise from modifications
to the Kerr geometry of each individual BH. Those finite-size effects can manifest themselves in
the QNMs of the remnant produced by the merger, but also in the GW phasing of the inspiral through
corrections to the GR spin-induced quadrupole and Love numbers. However, here we will not consider the latter,
instead, we will only study the impact of finite-size effects on the QNMs of the
remnant.

@@

The theory admits as a solution, the garden-variety Schwarzschild BH of GR.
This is not the case when rotation is included and the Kerr metric is not a
solution of the theory~\cite{Jackiw:2003pm}.
%
These rotating BHs support a scalar field which fall off as
$r^{-2}$ asymptotically, to which we can associate a scalar dipole
charge~\cite{Yunes:2009hc,Konno:2009kg} and
%
the leading-order modification to the GW phase enters at 2PN~\cite{Yagi:2011xp}.
%
Deviations from GR at this PN order are constrained with present GW observations~\cite{LIGOScientific:2021sio}
only at the level of $\sim 50\%$ (see the constraint on the 2PN parameter $\varphi_4$ in Fig.~6 of
Ref.~\cite{LIGOScientific:2021sio}). So far, analyses that used only the inspiral portion of
the BBH GW signals were not able to set meaningful bounds on such deviation at 2PN order~\cite{Nair:2019iur,Perkins:2021mhb}.
%
Nonetheless, the theory has been constrained in Ref.~\cite{Silva:2020acr}, which found
%
$\ell_{\rm dCS} \leqslant 8.5$~km,
%
by folding data from the X-ray observations of the isolated NS
PSR~J0030+0451~\cite{Lommen:2000yt,NANOGrav:2017wvv} by
NICER~\cite{Riley:2019yda,Miller:2019cac} and from the GW observation of the binary
NS GW170817~\cite{TheLIGOScientific:2017qsa,LIGOScientific:2018cki}, using
equation-of-state independent relations between the NS moment of inertia and tidal deformability~\cite{Yagi:2013bca,Yagi:2013awa,Gupta:2017vsl}.

@@

More specifically, we follow the notation of
Refs.~\cite{Cano:2020cao,Cano:2021myl} and consider up to dimension-eight
operators ($n=4$),
%
\begin{subequations}
\label{eq:action_mods}
\begin{align}
    L^{(6)} &= \lame R_{\mu\nu}{}^{\rho\sigma} R_{\rho\sigma}{}^{\gamma\delta} R_{\gamma \delta}{}^{\mu\nu}
    + \lamo R_{\mu\nu}{}^{\rho\sigma} R_{\rho\sigma}{}^{\gamma\delta} \tilde{R}_{\gamma \delta}{}^{\mu\nu}\,,
    \label{eq:s_6}
    \\
    L^{(8)} &= \varepsilon_{1} \mathcal{C}^2
    + \varepsilon_{2} \tilde{\mathcal{C}}^{2}
    + \varepsilon_{3} \mathcal{C} \tilde{\mathcal{C}}\,,
\label{eq:s_8}
\end{align}
\end{subequations}
%
where ${\cal C} = R_{\mu\nu\rho\sigma} R^{\mu\nu\rho\sigma}$,
$\tilde{\cal C} = R_{\mu\nu\gamma\delta} \epsilon^{\mu\nu}{}_{\rho\sigma} R^{\rho\sigma\gamma\delta}$,
and both $\lambda_{\rm o, e}$ and $\varepsilon_{i}$ (with $i=1, 2, 3$) are dimensionless parameters.
%
Due to the large number of free parameters in this theory, we focus on a subset of the parameter space.
%
In particular, we consider dimension-six and dimension-eight operators separately.
%
In addition, in the dimension-six case we further assume that $\lame = \lamo = 1$,
leaving us with $\ell_{\rm cEFT}$ as our single free parameter.
%
Similarly, in the dimension-eight case, we set $\varepsilon_{1} = 1$ and $\varepsilon_{2} = \varepsilon_{3} = 0$,
as done in Ref.~\cite{Sennett:2019bpc}.
%
This leaves us with $\ell_{\rm qEFT}$ as our single free parameter.

For the EFT of GR with dimension-eight operators, Ref.~\cite{Sennett:2019bpc} focused on the
orbital effects (i.e., instead of finite-size effects) and performed
Bayesian model selection using the two lowest-mass BBHs events of the second-observig run of
the LIGO-Virgo Collaboration, notably GW151226 and GW170608. They found that the
data disfavor the appearance of new physics on distance scales around $\ell_{\rm qEFT} \sim 150$ km.
